---
title: "Verasight Data Scientist Case Study – Task 2: Response Density Analysis"
author: "Jackson Zhao"
output: html_notebook
---

# Executive Summary: Response Density Analysis

## 🎯 Business Problem
**Density Analysis:** Measure whether survey responses are concentrated among a small number of heavy users or distributed evenly across many panelists. This helps understand user engagement patterns and potential bias in survey results.

## 📊 Final Results

### **Task 2.1 Answer:** 15.68% of all users account for 50% of survey responses
### **Task 2.2 Answer:** 28.66% of recent users (90 days) account for 50% of their responses  
### **Task 2.3 Answer:** Lorenz curve visualization showing concentration at all thresholds

---


## 📋 Data Processing Workflow

### **Step 1: Data Cleaning Strategy**
- **Remove missing IDs:** Can't attribute responses without user identifiers
- **Standardize formats:** Ensure consistent ID types for proper joins
- **Parse dates:** Enable accurate 90-day filtering

### **Step 2: User Response Counting**
- **Aggregate:** Count total responses per user ID
- **Rank:** Sort users by response count (descending)
- **Calculate:** Running totals and percentages

### **Step 3: Cumulative Analysis**
- **Running sum:** Cumulative responses from top users
- **Percentage calculation:** Convert to proportion of total responses
- **Threshold detection:** Find where cumulative share crosses target

### **Step 4: Lorenz Curve Construction**
- **X-axis:** Cumulative percentage of users (0% to 100%)
- **Y-axis:** Cumulative percentage of responses (0% to 100%)
- **Interpretation:** Closer to diagonal = more equal distribution

---

# Task 2.1: All Users Density Analysis

## Summary
Calculate minimum percentage of ALL Verasight users accounting for 50% of survey responses.

## Workflow
1. **Join datasets:** Link response database with user information
2. **Count responses:** Aggregate responses per user ID
3. **Sort & rank:** Order users by response count (highest first)
4. **Calculate cumulative:** Running totals of responses and user percentages
5. **Find threshold:** Identify where cumulative responses ≥ 50%

## Answer: **15.68% of users account for 50% of all survey responses**

---

# Task 2.2: Recent Users (90 Days) Density Analysis

## Summary
Calculate same density metric but restricted to users who registered within last 90 days.

## Workflow
1. **Define 90-day window:** Calculate cutoff date from most recent signup
2. **Filter recent users:** Subset to users registered within window
3. **Subset responses:** Only include responses from recent users
4. **Recalculate metrics:** Apply same cumulative analysis to recent cohort
5. **Compare results:** Analyze difference vs all-time users

## Answer: **28.66% of recent users account for 50% of responses from that cohort**

---

# Task 2.3: Visualization - Lorenz Curve Analysis

## Summary
Create detailed visualization showing density across multiple thresholds (10%, 20%, 30%, etc.).

## Workflow
1. **Generate thresholds:** Calculate user percentages for 10%-90% response shares
2. **Create Lorenz curve:** Plot cumulative users vs cumulative responses
3. **Add reference line:** Perfect equality diagonal for comparison
4. **Interpret results:** Distance from diagonal shows concentration level

## Answer: **Interactive Lorenz curve showing power-law distribution pattern**

**Key Insights:**
- **High concentration:** Small elite drives majority of responses
- **Power-law pattern:** Typical of user engagement platforms
- **Inequality measure:** Distance from diagonal indicates response concentration

---

# 📈 Business Insights & Conclusions

## Key Findings Summary

### **Concentration Patterns:**
1. **All Users:** 15.68% account for 50% of responses → High concentration
2. **Recent Users:** 28.66% account for 50% of responses → Even higher concentration  
3. **Early Threshold:** Only 2.07% of users account for 10% of responses

### **Strategic Implications:**
- **Survey bias risk:** Small user subset drives majority of feedback
- **Engagement opportunity:** Large portion of users underutilized  
- **Quality focus:** Heavy users require special attention for data quality
- **Growth potential:** Recent users show higher engagement rates

### **Technical Validation:**
- **Formula accuracy:** `min(which())` correctly identifies threshold crossings
- **Data integrity:** Proper handling of missing IDs ensures valid user attribution
- **Visualization clarity:** Lorenz curve effectively shows concentration patterns

## File Outputs Generated

**Output Structure:**
```
task2/outputs/
├── all_users_analysis.csv         # Detailed user-level analysis
├── recent_users_analysis.csv      # 90-day cohort analysis  
├── density_thresholds_summary.csv # Multiple threshold results
├── all_vs_recent_comparison.csv   # Comparative analysis
├── lorenz_curve_all_users.png     # Lorenz curve visualization
└── user_comparison.png            # All vs recent comparison plot
```

---

# 💻 Code Implementation

Below is the complete technical implementation of the analysis described above.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(plotly)
library(scales)

# Define relative paths from task2 folder
data_dir <- "../../data"  # Go up two levels to reach project root, then into data
output_dir <- "./outputs" # Create outputs folder within task2

# Create output folder if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Load datasets
full_db <- readRDS(file.path(data_dir, "full-response-db.rds"))
users <- readRDS(file.path(data_dir, "users.rds"))
```

```{r data-cleaning}
cat("=== DATA CLEANING & PREPARATION ===\n")

# Check missing values
cat("Missing IDs in response database:", sum(is.na(full_db$ID)), "rows\n")
cat("These must be removed because we cannot attribute responses to users without IDs\n\n")

# Remove rows with missing user ID (cannot be attributed to any user)
full_db_clean <- full_db %>%
  filter(!is.na(ID))

cat("After cleaning - usable responses:", nrow(full_db_clean), "\n")
```

```{r task2-1-implementation}
cat("=== TASK 2.1: ALL USERS ANALYSIS ===\n")

# Step 1: Count responses per user
user_counts <- full_db_clean %>%
  count(ID, name = "response_count")

# Step 2: Join with user data and calculate cumulative statistics
user_data <- user_counts %>%
  left_join(users, by = "ID") %>%
  arrange(desc(response_count)) %>% # Sort by highest responders first
  mutate(
    # Calculate running totals
    cumulative_responses = cumsum(response_count),
    total_responses = sum(response_count),
    cumulative_pct = cumulative_responses / total_responses,
    user_rank = row_number(),
    user_pct = user_rank / n() # Percentage of users represented
  )

# Step 3: Find minimum users for 50% threshold using min(which())
cutoff_row <- min(which(user_data$cumulative_pct >= 0.5))
pct_50_all <- user_data$user_pct[cutoff_row] * 100

cat(sprintf("ANSWER: %.2f%% of users account for 50%% of all responses\n", pct_50_all))

# Save detailed analysis
write_csv(user_data, file.path(output_dir, "all_users_analysis.csv"))
```

```{r task2-2-implementation}
cat("=== TASK 2.2: RECENT USERS (90 DAYS) ===\n")

# Step 1: Parse signup dates and define 90-day window
users$signup_date <- as.Date(users$signup_date, format = "%m/%d/%Y")
cutoff_date <- max(users$signup_date, na.rm = TRUE) - 90
cat("90-day cutoff date:", as.character(cutoff_date), "\n")

# Step 2: Filter to recent users only
recent_users <- users %>%
  filter(signup_date >= cutoff_date)
cat("Recent users found:", nrow(recent_users), "\n")

# Step 3: Subset response data for recent users
recent_data <- user_data %>%
  filter(ID %in% recent_users$ID) %>%
  arrange(desc(response_count)) %>%
  mutate(
    # Recalculate cumulative stats for recent cohort only
    cumulative_responses = cumsum(response_count),
    total_responses = sum(response_count),
    cumulative_pct = cumulative_responses / total_responses,
    user_rank = row_number(),
    user_pct = user_rank / n()
  )

# Step 4: Find 50% threshold for recent users
cutoff_row_recent <- min(which(recent_data$cumulative_pct >= 0.5))
pct_50_recent <- recent_data$user_pct[cutoff_row_recent] * 100

cat(sprintf("ANSWER: %.2f%% of recent users account for 50%% of their responses\n", pct_50_recent))

# Save recent users analysis
write_csv(recent_data, file.path(output_dir, "recent_users_analysis.csv"))

# Comparison insight
cat("\nCOMPARISON:\n")
cat("All users:", round(pct_50_all, 2), "%\n")
cat("Recent users:", round(pct_50_recent, 2), "%\n")
if(pct_50_recent < pct_50_all) {
  cat("Recent users show HIGHER concentration (more active heavy users)\n")
} else {
  cat("Recent users show LOWER concentration (more distributed engagement)\n")
}
```

```{r task2-3-implementation}
cat("=== TASK 2.3: MULTIPLE THRESHOLD ANALYSIS ===\n")

# Calculate user percentages needed for various response thresholds
response_thresholds <- c(0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90)

density_summary <- sapply(response_thresholds, function(thresh) {
  cutoff_index <- min(which(user_data$cumulative_pct >= thresh))
  user_data$user_pct[cutoff_index] * 100
})

names(density_summary) <- paste0(response_thresholds * 100, "%")

cat("User percentage needed for each response threshold:\n")
for (i in 1:length(density_summary)) {
  cat(sprintf("• %.2f%% of users → %s of responses\n", 
              density_summary[i], names(density_summary)[i]))
}

# Create summary table
summary_table <- data.frame(
  Response_Threshold = names(density_summary),
  User_Percentage_Needed = round(density_summary, 2)
)

write_csv(summary_table, file.path(output_dir, "density_thresholds_summary.csv"))
```

```{r lorenz-curve-visualization}
cat("=== LORENZ CURVE VISUALIZATION ===\n")

# Create Lorenz curve plot
lorenz_plot <- ggplot(user_data, aes(x = user_pct, y = cumulative_pct)) +
  geom_line(color = "darkblue", size = 1.2, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "gray50", alpha = 0.7) +
  annotate("text", x = 0.7, y = 0.3, 
           label = "Perfect Equality\n(45° line)", 
           color = "gray50", size = 3) +
  annotate("text", x = 0.3, y = 0.7, 
           label = "Actual Distribution\n(High Concentration)", 
           color = "darkblue", size = 3) +
  scale_x_continuous(labels = scales::percent, name = "Cumulative % of Users") +
  scale_y_continuous(labels = scales::percent, name = "Cumulative % of Responses") +
  labs(title = "Response Density: Lorenz Curve Analysis",
       subtitle = paste0("Distance from diagonal shows concentration level | ",
                        round(pct_50_all, 1), "% of users → 50% of responses")) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(size = 14, face = "bold"))

# Display interactive version
interactive_lorenz <- ggplotly(lorenz_plot, tooltip = c("x", "y"))
interactive_lorenz

# Save static version
ggsave(file.path(output_dir, "lorenz_curve_all_users.png"), lorenz_plot, 
       width = 10, height = 6, dpi = 300)

cat("Lorenz Curve Interpretation:\n")
cat("• Diagonal line = Perfect equality (everyone contributes equally)\n")
cat("• Curved line = Actual distribution (concentrated among few users)\n")
cat("• Larger area between lines = Higher concentration\n")
```

```{r comparison-all-vs-recent}
cat("=== COMPARISON: ALL vs RECENT USERS ===\n")

# Calculate thresholds for recent users
recent_summary <- sapply(response_thresholds, function(thresh) {
  if (max(recent_data$cumulative_pct) >= thresh) {
    cutoff_index <- min(which(recent_data$cumulative_pct >= thresh))
    recent_data$user_pct[cutoff_index] * 100
  } else {
    NA
  }
})

# Create comparison dataset
comparison_data <- data.frame(
  Response_Threshold = response_thresholds * 100,
  All_Users = density_summary,
  Recent_Users = recent_summary
) %>%
  pivot_longer(cols = c(All_Users, Recent_Users), 
               names_to = "User_Group", 
               values_to = "User_Percentage")

# Create comparison plot
comparison_plot <- ggplot(comparison_data, aes(x = Response_Threshold, y = User_Percentage, 
                                              color = User_Group, linetype = User_Group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(name = "Response Threshold (%)", breaks = seq(10, 90, 10)) +
  scale_y_continuous(name = "User Percentage Needed (%)") +
  scale_color_manual(values = c("All_Users" = "darkblue", "Recent_Users" = "darkred")) +
  labs(title = "Response Concentration: All Users vs Recent Users (90 Days)",
       subtitle = "Lower values indicate higher concentration among fewer users",
       color = "User Group", linetype = "User Group") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "bottom")

print(comparison_plot)
ggsave(file.path(output_dir, "user_comparison.png"), comparison_plot, 
       width = 10, height = 6, dpi = 300)

# Save comparison data
write_csv(comparison_data, file.path(output_dir, "all_vs_recent_comparison.csv"))
```

```{r final-summary}
cat("=== ANALYSIS COMPLETE ===\n")
cat("Generated files in", normalizePath(output_dir), ":\n")
output_files <- list.files(output_dir)
for (file in output_files) {
  cat("•", file, "\n")
}

cat("\nFinal Answers:\n")
cat("Task 2.1:", round(pct_50_all, 2), "% of all users → 50% of responses\n")
cat("Task 2.2:", round(pct_50_recent, 2), "% of recent users → 50% of their responses\n")
cat("Task 2.3: Lorenz curve visualization with multiple threshold analysis\n")
``` 